---
title: "R bootcamp exercise - Fisheries stakeholders survey"
author: "Daniel Viana and Jessica Zamborain Mason"
date: "October, 2022"
output:
  word_document:
    toc: yes
  html_document:
    toc: yes
    toc_float: yes
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We are going to work with data from a survey conducted with fisheries stakeholders in Brazil. Through an online survey, they were asked if they agree or diagree with the use of no-take areas for fisheries management. Respondents chose a response from 1 to 5, where: 1 - Agree, 2 - Partially agree, 3 - Neither agree or disagree, 4 - Partially disgree, 5 - Disagree.  

1) Read explore the data ("data_qualitative.csv"). How many variables does the dataset have? What information it contains? How many respondents?
2) Calculate the frequency distribution of responses. How many responses did we have for each option (1 to 5)?
3) Calculate the relative frequency and percentage distributions of responses. What percentage of responses was 1, 2, 3, 4 and 5?
4) Build a bar graph of the frequency distribution of responses.
5) Build a pie chart of percentage distributions distribution of responses.


##Answers
1) Read explore the data ("data_qualitative.csv"). How many variables does the dataset have? What information it contains? How many respondents?
```{r}
library(tidyverse)
dat <- read_csv("~/Rbootcamp_Madagascar/data/data_qualitative.csv")
nrow(dat)
```

2) Calculate the frequency distribution of responses. How many responses did we have for each option (1 to 5)?
```{r}
dat %>% 
  count(value)
```

3) Calculate the relative frequency and percentage distributions of responses. What percentage of responses was 1, 2, 3, 4 and 5?
```{r}
dat %>% 
  count(value) %>% 
  mutate(rel_freq = n/nrow(dat),
         perc = rel_freq*100)
```

4) Build a bar graph of the frequency distribution of responses.

```{r}
freq_dat = dat %>% 
  count(value) %>% 
  mutate(rel_freq = n/nrow(dat),
         perc = rel_freq*100,
         response = case_when(value == 1 ~ "Agree", 
                              value == 2 ~ "Partially agree",
                              value == 3 ~ "Neither agree or disagree", 
                              value == 4 ~ "Partially disagree", 
                              value == 5 ~ "Disagree"))

ggplot(data = freq_dat, aes(x = reorder(response, n), y = n)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(x = "", y = "Frequency") +
  theme_classic()
```
5) Build a pie chart of percentage distributions distribution of responses.
```{r}
freq_dat = dat %>% 
  count(value) %>% 
  mutate(rel_freq = n/nrow(dat),
         perc = rel_freq*100,
         response = case_when(value == 1 ~ "Agree", 
                              value == 2 ~ "Partially agree",
                              value == 3 ~ "Neither agree or disagree", 
                              value == 4 ~ "Partially disagree", 
                              value == 5 ~ "Disagree"),
         lab = paste(round(perc), "%", sep = ""),
         ypos = c(80, 52, 25, 3))

ggplot(data = freq_dat, aes(x = "", y = perc, fill = response)) +
  geom_bar(width=1, color="white", stat = "identity") +
  coord_polar("y", start=0) +
  theme_void() +
  labs(fill = "") +
  geom_text(aes(y = ypos, label = lab), color = "white", size=6) +
  scale_fill_brewer(palette="Set1")
```
